---
title: "Warm-up-mosquitoes-PATTERSON"
author: "Grant Patterson"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    df_print: paged
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

Fill in abstract at the end after we have finished the methods, results, discussion, conclusions and know what our data "says".

# BACKGROUND

The West Nile Virus (WNV) is a mosquito-born virus that proliferates in avian hosts and can be transferred to various different host, both human and animal. House finches populations in particular have been shown to have been shown to have a positive correlation with WNV-positive collected blood meals (Komer, 2003). The data used in this report was gathered in the Salt Lake UT area with the intent of analyzing the most common avian hosts of the virus in areas with high percentages of WNV-positive tests on mosquito blood meals.

To collect these blood meals, mosquitoes were caught in traps spread around Salt Lake and the surrounding area. Female mosquitoes were then separated from the rest and crushed to release the blood meal. Using Polymerase Chain Reactions (PCR) a specific sequence unique to vertebrates could then be targeted and replicated on mass. This produced many copies of the host’s gene that could then be sequenced to determine its identity. This information was then analyzed to find the hosts that most contributed to the proliferation of WNV.

```{r viremia, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# Manually transcribe duration (mean, lo, hi) from the last table column
bird_data <- data.frame(
  Bird = c("Canada Goose","Mallard", 
           "American Kestrel","Northern Bobwhite",
           "Japanese Quail","Ring-necked Pheasant",
           "American Coot","Killdeer",
           "Ring-billed Gull","Mourning Dove",
           "Rock Dove","Monk Parakeet",
           "Budgerigar","Great Horned Owl",
           "Northern Flicker","Blue Jay",
           "Black-billed Magpie","American Crow",
           "Fish Crow","American Robin",
           "European Starling","Red-winged Blackbird",
           "Common Grackle","House Finch","House Sparrow"),
  mean = c(4.0,4.0,4.5,4.0,1.3,3.7,4.0,4.5,5.5,3.7,3.2,2.7,1.7,6.0,4.0,
           4.0,5.0,3.8,5.0,4.5,3.2,3.0,3.3,6.0,4.5),
  lo   = c(3,4,4,3,0,3,4,4,4,3,3,1,0,6,3,
           3,5,3,4,4,3,3,3,5,2),
  hi   = c(5,4,5,5,4,4,4,5,7,4,4,4,4,6,5,
           5,5,5,7,5,4,3,4,7,6)
)

# selecting colors
cols <- c(rainbow(30)[c(10:29,1:5)])  # rainbow colors

# horizontal barplot
par(mar=c(5,12,2,2))  # wider left margin for names
bp <- barplot(bird_data$mean, horiz=TRUE, names.arg=bird_data$Bird,
              las=1, col=cols, xlab="Days of detectable viremia", xlim=c(0,7))

# add error bars to horizontal bar plot
arrows(bird_data$lo, bp, bird_data$hi, bp,
       angle=90, code=3, length=0.05, col="black", xpd=TRUE)
```

# STUDY QUESTION and HYPOTHESIS

## Questions  
Which bird species acts as an amplifying host of WNV in the Salt Lake City area?

## Hypothesis 
There is a strong correlation between house finch populations and WNV proliferation in a given area due to house finches being a good host for the virus.
  
## Prediction
If house finches are acting as important amplifying hosts, we predict that trapping locations where mosquitoes feed on house finches will also have higher rates of confirmed WNV in tested mosquito pools.

# METHODS

Fill in here, including overview of procedure and methods used for this project.

## Fill in first analysis
```{r first-analysis, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
## insert code for analysis part here - cut do to not being able to get files to function properly
```

## Fill in second analysis/plot
```{r second-analysis-or-plot, include=TRUE, eval=TRUE, warning=FALSE, fig.align='left', fig.width=4.75, fig.height=7}
# put code for plotting here
```

# DISCUSSION

## Interpretation - fill in analysis

## Interpretation - fill in analysis/plot

# CONCLUSION
The data indicates that there is a strong correlation between Housefinch blood meals and West Nile Virus spread in the Salt Lake Area

# REFERENCES     

1. Komar N, Langevin S, Hinten S, Nemeth N, Edwards E, Hettler D, Davis B, Bowen R, Bunning M. Experimental infection of North American birds with the New York 1999 strain of West Nile virus. Emerg Infect Dis. 2003 Mar;9(3):311-22. https://doi.org/10.3201/eid0903.020628

2. Google Gemini. Google, version Sep 2025. Used as a reference for functions such as plot() and to correct syntax errors. Accessed `r Sys.Date()`.